---
title: "Deforestation Detection"
author: "Brian Pondi, Jonathan Bahlmann"
date: "2/12/2021"
output: 
 md_document:
   variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* Observe structure: Intro, Methods, Results, Discussion, Conclusion

# Introduction

# Methods

The time series Landsat 8 satellite images used for this research were already provided in a state where cloud cover were already removed. The data covers the period 01-01-2013 to 31-12-2019. 
The Landsat 8 data covered row 001, paths 066 and 067.

The time series data was aggregated to a monthly and quarterly NDVI (Normalized Difference Vegetation Index) data where the median pixel values per row and column were selected. This resulted to 12 NDVI images per year in the former approach and 4 NDVI images per year in the latter approach


Prediction of missing values in satellite data are carried out using gapfill package in R . The gapfill approach was designed to carry out predictions on satellite data that  were recorded at equally spaced points of time.  Based on Gerber et.al 2016 , they applied the algorithm to MODIS NDVI data with cloud cover scenarios of 50% missing data. The method was further compared to Gapfill-Python and TIMESAT  and it provided the most accurate prediction in terms of RMSE.

Gapfill was appealing to this research because it's capable of handling large spatio-temporal data, it’s user friendly and capability tailored to  specific features of different satellite images. The predictions of the missing values are based on a subset-predict procedure, i.e., each missing value is predicted separately by (1) selecting subsets of the data that are in a neighborhood around the missing point and (2) predicting the missing value based on the subset(Gerber, 2016).

In this research we also explored to tailor gapfill by customizing the iMax parameter which is the maximum number of iterations until NA is  returned as predicted value (Gerber, 2016). The research  compares the iMax parameter at value 5 (five)  for  five iterations versus undefined which results to infinite iterations by default.


BFAST(Breaks For Additive Season and Trend)


## Input Data and Preparations

* Package description
* PRODES data sorted by years can be found here: [PRODES yearly deforestation](http://terrabrasilis.dpi.inpe.br/download/dataset/legal-amz-prodes/vector/yearly_deforestation.zip)

As seen in "final.Rmd". The subdirectory `L8cube_subregion` contains a NDVI time series as single `.tif` files, a file per acquisition, as input data.
```{r load-data, message=FALSE, warning=FALSE}
library(stars)
library(gapfill)
library(bfast)
library(zoo)
library(raster)
subdir = "landsat_monthly"
subdir = "landsat_quarterly"
f = paste0(subdir, "/", list.files(subdir))
st = merge(read_stars(f)) # make stars object
plot(st)

# load PRODES data
# prod <- read_sf("./yearly_deforestation/yearly_deforestation.shp")
# prod_3857 <- st_make_valid(st_transform(prod, crs = st_crs(st)))
# prod_crop <- st_crop(prod_3857, st) # clip
# write_sf(prod_crop, "./yearly_deforestation/PRODES_cropped.shp", overwrite = TRUE)
prod <- read_sf("./yearly_deforestation/PRODES_cropped.shp")
# prod <- prod[prod$YEAR < 2019,]
```

```{r plot-AOI, warning=FALSE}
# whats in it?
# plot(st)
plot(prod["YEAR"])
```

## Prepare for Gapfill

`Gapfill` documentation tells us that as input, a 4-dimensional numeric array is needed, with dimensions x, y, seasonal index (doy), year.
```{r create-gapfill-input}
# get pixels of whole dataset
imgdata <- c(st[,,,][[1]])

# make labels
xlab <- seq(from = attr(st, "dimensions")[[1]]$offset, by = attr(st, "dimensions")[[1]]$delta, length.out = attr(st, "dimensions")[[1]]$to)
ylab <- seq(from = attr(st, "dimensions")[[2]]$offset, by = attr(st, "dimensions")[[2]]$delta, length.out = attr(st, "dimensions")[[2]]$to)
doi <- c(1, 32, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335)
doi <- c(1, 91, 182, 274)
years <- seq(2013,2019,1)

# make array, transpose
h <- array(imgdata, dim = c(140, 140, 4, 7), dimnames = list(xlab, ylab, doi, years))
h <- aperm(h, c(2,1,3,4))

# all
Image(h[,,2,3:4])
# original stars
plot(st[,,,c(5, 17)])
```

## Gapfill

```{r do-gapfill, warning=FALSE}
# d <- Gapfill(h, iMax = 5)
# saveRDS(d, "./iMax5_140_gapfilled_quarterly.rds")
gf_monthly <- readRDS("monthly_iMax5_140_gapfilled.rds")
Image(gf_monthly$fill)
gf_quarterly <- readRDS("quarterly_iMax5_140_gapfilled.rds")
Image(gf_quarterly$fill)
```

## BFAST Test
```{r bfast-test, eval=FALSE}
# ts of a pixel
x <- as.vector(gf_monthly$fill[15,20,,])
# time must be given scaled to 1. -> monthly -> 1/12 = .08333333
# zoo must also be used in between, exactly as in ?bfastmonitor example
y <- as.ts(zoo(x, seq(2013, by = .08333333, length.out = 84))) 
bf <- bfastmonitor(y, start = 2019, order = 3, verbose = TRUE) # 2019,6 works but monitoring period is then off chart
plot(bf)

x1 <- as.vector(gf_quarterly$fill[15,20,,])
y1 <- as.ts(zoo(x1, seq(2013, by = .24999999, length.out = 28))) 
bf1 <- bfastmonitor(y1, start = 2019, order = 2, verbose = TRUE) # 2019,6 works but monitoring period is then off chart
plot(bf1)
```

## Calculate BFAST Over AOI

```{r bfast-aoi, warning=FALSE, eval=FALSE}
bfast_on_tile <- function(gapfill_matrix, by, ts, order) {
  dims <- dim(gapfill_matrix)
  result <- matrix(rep(FALSE, dims[1]*dims[2]), ncol = dims[1])
  for (i in 1:dims[1]) { # looping through x
    for (j in 1:dims[2]) { # loops through y
      raw_px_ts <- as.vector(gapfill_matrix[i,j,,])
      px_ts_obj <- as.ts(zoo(raw_px_ts, seq(2013, by = by, length.out = ts)))
      bfm_obj <- bfastmonitor(px_ts_obj, start = 2019, order = order)
      brkpoint <- bfm_obj$breakpoint
      if(!is.na(brkpoint)) {
        result[i,j] <- TRUE
      } else {
        # FALSE
      }
    }
  }
  return(result)
}

bfast_monthly3 <- bfast_on_tile(gf_monthly$fill, by = .08333333, ts = 84, order = 3)
bfast_monthly2 <- bfast_on_tile(gf_monthly$fill, by = .08333333, ts = 84, order = 2)
bfast_quarter2 <- bfast_on_tile(gf_quarterly$fill, by = 0.25, ts = 28, order = 2)
saveRDS(bfast_monthly3, "bfast_monthly3.rds")
saveRDS(bfast_monthly2, "bfast_monthly2.rds")
saveRDS(bfast_quarter2, "bfast_quarter2.rds")
# warning: too few observations in history period
```

```{r load-bfast-tiles}
# bfast_monthly3 <- readRDS("bfast_monthly3.rds")
bfast_monthly2 <- readRDS("bfast_monthly2.rds")
bfast_quarter2 <- readRDS("bfast_quarter2.rds")
```

```{r plot-bfast-aoi, warning=FALSE}
# rasterize PRODES data
ras <- rasterize(prod, as(st[,,,5], "Raster"), "YEAR")
prodes <- aperm(matrix(ras[], ncol = 140), c(2,1))
prodes[prodes < 2019] <- FALSE
prodes[prodes == 2019] <- TRUE
prodes[is.na(prodes)] <- FALSE

# to mask out previous deforestation
prodes_prev <- aperm(matrix(ras[], ncol = 140), c(2,1))
prodes_prev[prodes_prev < 2019] <- TRUE
prodes_prev[prodes_prev == 2019] <- FALSE
prodes_prev[is.na(prodes_prev)] <- FALSE

bfast_monthly2_prev <- bfast_monthly2
bfast_monthly2_prev[prodes_prev == 1] <- NA

bfast_quarter2_prev <- bfast_quarter2
bfast_quarter2_prev[prodes_prev == 1] <- NA
```

```{r aoi-tables}
table1 <- addmargins(table(bfast_monthly2, prodes))
table2 <- addmargins(table(bfast_quarter2, prodes))

accuracies <- function(table1) {
  # overall accuracy
  P0 <- (table1[1] + table1[5]) / table1[9]
  # error of omission, wrong in respect to reference data
  # table1[2] / table1[3] # FALSE
  # table1[4] / table1[6] # TRUE
  # error of commission, wrong in respect to classified data
  # table1[4] / table1[7] # FALSE
  # table1[2] / table1[8] # TRUE
  # producer's accuracy, Probability of classifying a pixel correctly
  pa_f <- table1[1] / table1[3] # FALSE
  pa_t <- table1[5] / table1[6] # TRUE
  # user's accuracy, Probability of a pixel being the classified type
  ua_f <- table1[1] / table1[7] # FALSE
  ua_t <- table1[5] / table1[8] # TRUE
  # kappa
  # chance that both TRUE / FALSE randomly
  tr <- (table1[8] / table1[9]) * (table1[6] / table1[9])
  fr <- (table1[7] / table1[9]) * (table1[3] / table1[9])
  Pe <- tr + fr
  kappa <- (P0 - Pe) / (1 - Pe)
  
  return(list("Overall Accuracy" = P0*100, "Prod. Acc. FALSE" = pa_f, "Prod. Acc. TRUE" = pa_t, "User's Acc. FALSE" = ua_f, "User's Acc. TRUE" = ua_t, "Kappa" = kappa))
}
```

# Results

```{r results, fig.show="hold", out.width="33%"}
Image(bfast_monthly2) + ggtitle("Monthly Data") + theme(plot.title = element_text(size=22))
Image(bfast_quarter2) + ggtitle("Quarterly Data") + theme(plot.title = element_text(size=22))
Image(prodes) + ggtitle("PRODES Data") + theme(plot.title = element_text(size=22))

table1
table2

array(c(accuracies(table1), accuracies(table2)), dim = c(6,2), dimnames = list(c("Overall Accuracy", "Prod. Acc. FALSE", "Prod. Acc. TRUE", "User's Acc. FALSE", "User's Acc. TRUE", "Kappa"), c("monthly", "quarterly")))

# remove already deforested areas
ras[] <- bfast_monthly2
plot(ras)
plot(st_geometry(prod), add = TRUE)

ras[] <- bfast_quarter2
plot(ras)
plot(st_geometry(prod), add = TRUE)


array(c(accuracies(addmargins(table(bfast_monthly2_prev, prodes))), accuracies(addmargins(table(bfast_quarter2_prev, prodes)))), dim = c(6,2), dimnames = list(c("Overall Accuracy", "Prod. Acc. FALSE", "Prod. Acc. TRUE", "User's Acc. FALSE", "User's Acc. TRUE", "Kappa"), c("monthly", "quarterly")))
```

# Discussion

# Conclusion

# References
Gerber F, Furrer R, Schaepman-Strub G, de Jong R, Schaepman ME (2016) Predicting missing values in spatio-temporal satellite data.

Verbesselt, J., Hyndman, R., Newnham, G., & Culvenor, D. (2010). Detecting trend and seasonal changes in satellite image time series. 

Verbesselt, J., Zeileis, A., & Herold, M. (2013). Near real-time disturbance detection using satellite image time series.

