---
title: "Deforestation Detection"
author: "Brian Pondi, Jonathan Bahlmann"
date: "2/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* Observe structure: Intro, Methods, Results, Discussion, Conclusion

# Introduction

# Methods

Prediction of missing values in satellite data ere carried out using gapfill package in R . The gapfill approach was designed to carry out predictions on satellite data that  were recorded at equally spaced points of time.  Based on Gerber et.al 2016 , they applied the algorithm to MODIS NDVI data with cloud cover scenarios of 50% missing data. The method was further compared to Gapfill-Python and TIMESAT  and it provided the most accurate prediction in terms of RMSE.

Gapfill was appealing to this research because it's capable of handling large spatio-temporal data, it’s user friendly and capability tailored to  specific features of different satellite images. The predictions of the missing values are based on a subset-predict procedure, i.e., each missing value is predicted separately by (1) selecting subsets of the data that are in a neighborhood around the missing point and (2) predicting the missing value based on the subset(Gerber, 2016).

In this research we also explored to tailor gapfill by customizing the iMax parameter which is the maximum number of iterations until NA is  returned as predicted value (Gerber, 2016). The research  compares the iMax parameter at value 5 (five)  for  five iterations versus undefined which results to infinite iterations by default.

BFAST(Breaks For Additive Season and Trend)


## Input Data and Preparations

* Package description
* PRODES data sorted by years can be found here: [PRODES yearly deforestation](http://terrabrasilis.dpi.inpe.br/download/dataset/legal-amz-prodes/vector/yearly_deforestation.zip)

As seen in "final.Rmd". The subdirectory `L8cube_subregion` contains a NDVI time series as single `.tif` files, a file per acquisition, as input data.
```{r load-data, message=FALSE, warning=FALSE}
library(stars)
library(gapfill)
library(bfast)
library(zoo)
library(raster)
subdir = "landsat_monthly"
subdir = "landsat_quarterly"
f = paste0(subdir, "/", list.files(subdir))
st = merge(read_stars(f)) # make stars object
plot(st)

# load PRODES data
# prod <- read_sf("./yearly_deforestation/yearly_deforestation.shp")
# prod_3857 <- st_make_valid(st_transform(prod, crs = st_crs(st)))
# prod_crop <- st_crop(prod_3857, st) # clip
# write_sf(prod_crop, "./yearly_deforestation/PRODES_cropped.shp", overwrite = TRUE)
prod <- read_sf("./yearly_deforestation/PRODES_cropped.shp")
# prod <- prod[prod$YEAR < 2019,]
```

```{r plot-AOI, warning=FALSE}
# whats in it?
# plot(st)
plot(as(st[,,,6], "Raster"))
plot(prod["YEAR"], add = TRUE)
plot(prod["YEAR"])
```

## Prepare for Gapfill

`Gapfill` documentation tells us that as input, a 4-dimensional numeric array is needed, with dimensions x, y, seasonal index (doy), year.
```{r create-gapfill-input}
# get pixels of whole dataset
imgdata <- c(st[,,,][[1]])

# make labels
xlab <- seq(from = attr(st, "dimensions")[[1]]$offset, by = attr(st, "dimensions")[[1]]$delta, length.out = attr(st, "dimensions")[[1]]$to)
ylab <- seq(from = attr(st, "dimensions")[[2]]$offset, by = attr(st, "dimensions")[[2]]$delta, length.out = attr(st, "dimensions")[[2]]$to)
doi <- c(1, 32, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335)
doi <- c(1, 91, 182, 274)
years <- seq(2013,2019,1)

# make array, transpose
h <- array(imgdata, dim = c(140, 140, 4, 7), dimnames = list(xlab, ylab, doi, years))
h <- aperm(h, c(2,1,3,4))

# all
Image(h[,,2,3:4])
# original stars
plot(st[,,,c(5, 17)])
```

## Gapfill

```{r do-gapfill, warning=FALSE}
# d <- Gapfill(h, iMax = 5)
# saveRDS(d, "./iMax5_140_gapfilled_quarterly.rds")
gf_monthly <- readRDS("monthly_iMax5_140_gapfilled.rds")
Image(gf_monthly$fill)
gf_quarterly <- readRDS("quarterly_iMax5_140_gapfilled.rds")
Image(gf_quarterly$fill)
```

## BFAST Test
```{r bfast-test, eval=FALSE}
# ts of a pixel
x <- as.vector(gf_monthly$fill[15,20,,])
# time must be given scaled to 1. -> monthly -> 1/12 = .08333333
# zoo must also be used in between, exactly as in ?bfastmonitor example
y <- as.ts(zoo(x, seq(2013, by = .08333333, length.out = 84))) 
bf <- bfastmonitor(y, start = 2019, order = 3, verbose = TRUE) # 2019,6 works but monitoring period is then off chart
plot(bf)

x1 <- as.vector(gf_quarterly$fill[15,20,,])
y1 <- as.ts(zoo(x1, seq(2013, by = .24999999, length.out = 28))) 
bf1 <- bfastmonitor(y1, start = 2019, order = 2, verbose = TRUE) # 2019,6 works but monitoring period is then off chart
plot(bf1)
```

## Calculate BFAST Over AOI

```{r bfast-aoi, warning=FALSE}
bfast_on_tile <- function(gapfill_matrix, by, ts, order) {
  dims <- dim(gapfill_matrix)
  result <- matrix(rep(FALSE, dims[1]*dims[2]), ncol = dims[1])
  for (i in 1:dims[1]) { # looping through x
    for (j in 1:dims[2]) { # loops through y
      raw_px_ts <- as.vector(gapfill_matrix[i,j,,])
      px_ts_obj <- as.ts(zoo(raw_px_ts, seq(2013, by = by, length.out = ts)))
      bfm_obj <- bfastmonitor(px_ts_obj, start = 2019, order = order)
      brkpoint <- bfm_obj$breakpoint
      if(!is.na(brkpoint)) {
        result[i,j] <- TRUE
      } else {
        # FALSE
      }
    }
  }
  return(result)
}

bfast_monthly <- bfast_on_tile(gf_monthly$fill, by = .08333333, ts = 84, order = 3)
bfast_quarter <- bfast_on_tile(gf_quarterly$fill, by = 0.25, ts = 28, order = 2)
# warning: too few observations in history period
```

```{r plot-bfast-aoi, warning=FALSE}
# plot(prod["YEAR"])
ras <- rasterize(prod, as(st[,,,5], "Raster"), "YEAR")
ras.m <- aperm(matrix(ras[], ncol = 140), c(2,1))
ras.m[ras.m < 2019] <- FALSE
ras.m[ras.m == 2019] <- TRUE
ras.m[is.na(ras.m)] <- FALSE

Image(bfast_monthly)
Image(bfast_quarter)
Image(ras.m)

table(bfast_monthly, ras.m)
table(bfast_quarter, ras.m)
```

## old code
```{r, eval=FALSE}
# single pixel
defo_test <- st[st_geometry(prod[4,])]
defo_test_ts <- as.vector(defo_test[,3,3,][[1]])
test_ts <- as.ts(zoo(defo_test_ts, seq(2013, by = .08333333, length.out = 84))) 
test <- bfastmonitor(test_ts, start = c(2018, 1))
plot(test)

# mean of this polygon
mean_test <- st_apply(st[st_geometry(prod[2,])], "attributes", mean,  na.rm = TRUE) # aggregate
defo_test_ts <- mean_test[][[1]]
test_ts <- as.ts(zoo(defo_test_ts, seq(2013, by = .08333333, length.out = 84))) 
test <- bfastmonitor(test_ts, start = c(2018, 1))
plot(test)

# calculate breaks for every single pixel
vec <- c(0)
count <- 1
for (i in 1:5) {
  for (j in 1:12) {
    ts <- defo_test[,i,j,][[1]]
    test_ts <- as.ts(zoo(ts, seq(2013, by = .08333333, length.out = 84)))
    
    if (length(unique(test_ts)) > 1) {
      test_bf <- bfastmonitor(test_ts, start = c(2018, 1))
      vec[count] <- test_bf$breakpoint
    } else {
        # vec[count] <- 99
    }
    count <- count + 1
  }
}
vec
```

# Results

# Discussion

# Conclusion

# Reference
Gerber et.al (2016). Predicting missing values in spatio-temporal satellite data.

Verbesselt, J., Hyndman, R., Newnham, G., & Culvenor, D. (2010). Detecting trend and seasonal changes in satellite image time series. Remote Sensing of Environment, 114, 106-115. 

