---
title: "Deforestation Detection"
author: "Brian Pondi, Jonathan Bahlmann"
date: "2/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* Observe structure: Intro, Methods, Results, Discussion, Conclusion

# Load Data
As seen in "final.Rmd". The subdirectory `L8cube_subregion` contains a NDVI time series as single `.tif` files, a file per acquisition, as input data.
```{r, warning=FALSE, message=FALSE}
library(stars)
library(gapfill)
subdir = "landsat_monthly"
f = paste0(subdir, "/", list.files(subdir))
st = merge(read_stars(f))
plot(st)
```

TODO: Introduce Gapfill, why did we chose this?

`Gapfill` documentation tells us that as input, a 4-dimensional numeric array is needed, with dimensions x, y, seasonal index (doy), year.
```{r}
# make array
helper <- c(st[,,,1][[1]])
for (i in 2:72) {
  helper <- c(helper, st[,,,i][[1]])
}
xlab <- seq(from = attr(st, "dimensions")[[1]]$offset, by = attr(st, "dimensions")[[1]]$delta, length.out = attr(st, "dimensions")[[1]]$to)
ylab <- seq(from = attr(st, "dimensions")[[2]]$offset, by = attr(st, "dimensions")[[2]]$delta, length.out = attr(st, "dimensions")[[2]]$to)
doi <- c(1, 32, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335)
years <- seq(2013,2018,1)

m <- array(helper, dim = c(540, 501, 12, 6), dimnames = list(xlab, ylab, doi, years))
# Image(m) # to plot, takes time
```

PRODES data sorted by years can be found here: [PRODES yearly deforestation](http://terrabrasilis.dpi.inpe.br/download/dataset/legal-amz-prodes/vector/yearly_deforestation.zip)

Next steps: cut to AOI etc.
```{r}
prod <- read_sf("./yearly_deforestation/yearly_deforestation.shp")
```